DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-05-04 manuscript/manuscript.tex
add some more reviews

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[english,man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
            pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\DeclareUnicodeCharacter{0301}{*************************************}

@@

\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=english]{babel}
\else
  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\fi

@@

Chronograms --phylogenies with branch lengths proportional to time-- represent key data on evolutionary time frame for the study of natural processes in many areas of biological research.
Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
The main DateLife workflow starts with one or more scientific taxon names provided by a user. Names are processed and standardized to a unified taxonomy. Then, DateLife runs a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extracts all chronograms that contain at least two queried taxon names, along with their metadata.
Finally, node ages from matching chronograms are mapped upon a chosen tree topology using the congruification algorithm. Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates appear generally correlated to those from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
Software such as DateLife can help increase awareness of the existing variation in alternative hypothesis of evolutionary time frame for the same organisms, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

Word count: 6111

@@

Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, \& others, 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).

Building a chronogram is not an easy task.
It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree. All of this entails specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.

Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis, research, teaching, and science communication and policy.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a computer-readable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.

@@

DateLife's core software applications are implemented in the R package \texttt{datelife}. Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from other biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain, 2018),
geiger (Pennell et al., 2014),
msa (Bodenhofer, Bonatesta, Horejš-Kainrath, \& Hochreiter, 2015),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

@@

DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013). TNRS also allows to correctly choose between homonyms, by considering other taxa provided as input to infer the taxonomic context of the homonym. DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy (OTT, Open Tree Of Life et al., 2016; Rees \& Cranston, 2017) as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis. Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Non-marine Genera (IRMNG) database (Rees et al., 2017).

Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains). If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag. If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query. In this case, subspecific variants are excluded.
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
Species and subspecific variant names are processed and searched as provided by the user.
The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the input names standardized to a taxonomy of choice (OTT by default), the corresponding OTT id numbers, and the topology of an input tree, if one was provided.

@@

At the time of writing of this manuscript
(Feb 21, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}). As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} R package, which is currently manually updated as new chronogram data is added to Phylesystem. The updated database is assigned a new version number, followed by a package release on CRAN.
Users can directly implement \texttt{datelife} functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or \enquote{phylo} objects.
\item
  The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
\end{enumerate}

@@

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009).
If the source chronograms do not form a grove, the supertree reconstruction will fail.
In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.

@@

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo et al., 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), and treePL (Smith \& O'Meara, 2012).

@@

\textbf{\emph{With no branch lengths.--}}
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008), which has proven useful for ecological analyses that require age data (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.

In the absence of genetic data, we consider that BLADJ is the most agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution. It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018). To do so, DateLife implements MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.

\textbf{\emph{Dating a tree with branch lengths.--}}
Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain markers for input taxa.
By default, mined genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015). Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older nodes (Forest et al., 2005). The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
The BOLD workflow is currently only available on DateLife's R package and not on the web application.

@@

Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R \enquote{phylo} format. Input and summary chronograms can be visualized and compared graphically, and users can construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\DeclareUnicodeCharacter{0301}{*************************************}

@@

\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}

% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Chronograms --phylogenies with branch lengths proportional to time-- represent key data on timing of evolutionary events for the study of natural processes in many areas of biological research.
Chronograms also provide valuable information that can be used for education, science communication, and conservation policy decisions.
Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
The main DateLife workflow starts with one or more scientific taxon names provided by a user. Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
Finally, node ages from matching chronograms are mapped to corresponding nodes from a chosen tree topology using the congruification algorithm. Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
Results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
We encourage the use of DateLife to help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

Word count: 6380

@@

Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, et al., 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).

Building a chronogram is not an easy task.
It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree. All of this entails specialized biological training, taxonomic domain knowledge, and a significant amount of research time, computational resources and funding.

Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org. It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data, also in a computer-readable format (Stoltzfus et al., 2013); and methods to summarize and compare the data.

@@

DateLife's core software applications are implemented in the R package \texttt{datelife},
and relies on functionalities from other biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Pennell et al., 2014),
msa (Bodenhofer, Bonatesta, Horejš-Kainrath, \& Hochreiter, 2015),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

@@

DateLife starts by processing an input consisting of the scientific name of at least one taxon. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013). TNRS also allows to correctly choose between homonyms, by considering other taxa provided as input to infer the taxonomic context of the homonym. DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy (OTT, Open Tree Of Life et al., 2016; Rees \& Cranston, 2017) as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis. Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Non-marine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).

Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains). If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query. In this case, subspecific variants are excluded.
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
Species and subspecific variant names are processed and searched as provided by the user.
The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the input names standardized to a taxonomy of choice (OTT by default), the corresponding OTT id numbers, and the topology of an input tree, if one was provided.

@@

At the time of writing of this manuscript
(May 04, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}). As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} R package, which is currently manually updated as new chronogram data is added to Phylesystem. The updated database is assigned a new version number, followed by a package release on CRAN.
Users can directly run \texttt{datelife} functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across source chronograms.
\item
  The source chronogram(s) with the most input taxon names.
\item
  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
\end{enumerate}

@@

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009).
If the source chronograms do not form a grove, the supertree reconstruction will fail.
In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.

@@

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo et al. (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), and treePL (Smith \& O'Meara, 2012).

@@

\textbf{\emph{With no branch lengths.--}}
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster {[}BLADJ; Webb et al. (2008); Webb and Donoghue (2005){]} algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008).
This approach has has proven useful for ecological analyses that require age data (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.

In the absence of genetic data, BLADJ is a very agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution. It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018). To do so, DateLife implements MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.

\textbf{\emph{Dating a tree with branch lengths.--}}
Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such trees requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.
By default, BOLD genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015). Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older nodes (Forest et al., 2005). The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.

@@

Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R ``phylo'' format. Input and summary chronograms can be visualized and compared graphically, and users can construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
